Genome-wide changes in expression during inflorescence architecture development

### 2015-08-06 ms.Rmd
final commit before moving to simpler struct and pipeline with ruffus

---
title: "Genome-wide changes in expression during inflorescence architecture development
  in rice"
csl: csl/the-plant-journal.csl
output:
  pdf_document:
    fig_caption: yes
    keep_md: yes
    keep_tex: yes
    number_sections: yes
  word_document:
    fig_caption: yes
  BiocStyle::html_document:
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    keep_md: yes
    number_sections: yes
bibliography: bib/test.bib
---

@@

To recover common expression patterns in the meristem samples, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do], and the standardised, VST-transformed read counts were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Futschik:2005vo]. To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between `cmin` and `cmax` (@SI). A cluster number of `{r} c` was chosen,

@@

## Patterns of expression of transcription families
+ **Figure**: GSEA of transcription factor families

## MADS genes: comparison of expression & homology in rice, tomato, arabidopsis
+ **Figure**: Similarity tree of MADS proteins with expression values and dN/dS values for each one.

>>>>

---
title: Genome-wide changes in expression during inflorescence architecture development
  in rice
csl: csl/the-plant-journal.csl
header-includes: \usepackage{graphicx}
output:
  word_document:
    fig_caption: yes
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
  BiocStyle::html_document:
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    keep_md: yes
    number_sections: yes
bibliography: bib/test.bib
---

```{r Setup, include=FALSE, results="hide", warning=FALSE}

library(data.table)
library(ggplot2)
extrafont::loadfonts()

knitr::opts_chunk$set(fig.path='figures/',
               cache.path='cache/',
               dev=c("cairo_pdf", "svg", "png"),
               fig.width=6.614,
               fig.height=6.614,
               fig.align = 'center',
               dpi=300,
               fig.show='hold',
               cache=TRUE,
               echo=FALSE,
               results="hide",
               message=FALSE,
               warning=FALSE)

# https://gist.githubusercontent.com/rmflight/3858973/raw/591ee603ce0996ac407902f7dd59520b35b3702a/tableFigureFuncs.r
pasteLabel <- function(preText, inObj, objName, insLink=TRUE){
	objNum <- inObj[objName]

	useText <- paste(preText, objNum, sep=" ")
	if (insLink){
		useText <- paste("[", useText, "](#", objName, ")", sep="")
	}
	useText
}

# this increments the counter, and gives a name to the number so we can reference it later
incCount <- function(inObj, useName){
	nObj <- length(inObj)
	useNum <- max(inObj) + 1
	inObj <- c(inObj, useNum)
	names(inObj)[nObj+1] <- useName
	inObj
}
figCount <- c("_"=0)
tableCount <- c("_"=0)

tableCat <- function(inFrame){
	outText <- paste(names(inFrame), collapse=" | ")
	outText <- c(outText, paste(rep("---", ncol(inFrame)), collapse=" | "))
	invisible(apply(inFrame, 1, function(inRow){
		outText <<- c(outText, paste(inRow, collapse=" | "))
	}))
	return(outText)
}
```

@@

```{r mfuzz}
expressionMatrix <- readRDS('output/cutadapt-2015-07-29/STAR-2015-07-29/DESeq2-2015-08-04/mfuzz-2015-08-05/expressionMatrix.Rds')
c1 <- readRDS('output/cutadapt-2015-07-29/STAR-2015-07-29/DESeq2-2015-08-04/mfuzz-2015-08-05/c1.Rds')
memCutoff <- 0.5

# get clusters and membership
cluster <- data.table(id = names(c1$cluster), Cluster = c1$cluster, Membership = apply(c1$membership, 1, max), key = "id")

# get standardised VST counts
exprs <- data.table(Biobase::exprs(expressionMatrix), keep.rownames = TRUE, key = "rn")

# set up cluster labels
numGenes <- cluster[Membership > 0.5, .(number = length(id)), by = Cluster]
setkey(numGenes, 'Cluster')
clustLabels <- numGenes[, paste0("Cluster ", Cluster, "\n(", number, " genes)")]

# merge clusters with expression values
plotData.wide <- cluster[exprs]

# make long
plotData <- reshape2::melt(plotData.wide[ Membership > memCutoff ], id.vars = c('id', 'Cluster', 'Membership'), variable.name = 'Stage', value.name = "Normalised, transformed read counts")

# add cluster labels
plotData[, Cluster := plyr::mapvalues(as.factor(Cluster), seq(1, length(unique(Cluster))), clustLabels)]

# fix stage label
plotData[, Stage := plyr::mapvalues(Stage, "ePBM.SBM", "ePBM/SBM")]

# set up heatscale
heatscale <- RColorBrewer::brewer.pal(n = 6, name = "YlOrRd")

# add centres
centres.wide <- data.table(c1$centers, Cluster = clustLabels)
centres <- reshape2::melt(centres.wide, id.vars = 'Cluster', variable.name = 'Stage', value.name = "Normalised, transformed read counts")
centres[, Stage := plyr::mapvalues(Stage, "ePBM.SBM", "ePBM/SBM")]

f_mfuzzClusters <- ggplot(plotData, aes(x = Stage, y = `Normalised, transformed read counts`,
                     colour = Membership, group = id)) +
  theme_minimal(base_size = 8, base_family = "Helvetica") +
  xlab(NULL) +
  scale_colour_gradientn(colours = heatscale, limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  geom_line() +
  geom_line(data = centres, mapping = aes(group = 1), colour = "black", alpha = 0.5) +
  facet_wrap(~ Cluster)

figCount <- incCount(figCount, "f_mfuzzClusters")
```

To recover common expression patterns in the meristem samples, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do], and the standardised, VST-transformed read counts were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Futschik:2005vo]. To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between `cmin` and `cmax` (@SI). A cluster number of `{r} c` was chosen (**`r I(pasteLabel("Figure", figCount, "f_mfuzzClusters"))`**).

<a id="f_mfuzzClusters"></a>
```{r f_mfuzzClusters, fig.cap="Fuzzy \\emph{c}-means clustering of normalised, transformed read counts"}
print(f_mfuzzClusters)
```

@@

## Patterns of expression of transcription families

+ **Figure**: GSEA of transcription factor families

## MADS genes: comparison of expression & homology in rice, tomato, arabidopsis

+ **Figure**: Similarity tree of MADS proteins with expression values and dN/dS values for each one.
